KIN: A method to infer relatedness from low-coverage ancient DNA

### 2022-04-07 paper_overleaf/main.tex

\title{KIN: Method to infer relatedness using low coverage ancient DNA}
\author{Divya Ratan Popli, Stéphane Peyrégne, Benjamin M. Peter}
\date{5 August 2021}
\linenumbers

@@

\noindent Knowledge of familial relationships is an important part of several research fields. Knowing genetic kinship of ancient specimens can lead to interesting insights into the culture and social hierarchy of ancient populations, and is also relevant for downstream genetic analyses. However, estimating relatedness from ancient DNA can be difficult since DNA sequences extracted from ancient remains may have low coverage, ascertainment bias, and contamination from modern human DNA. In addition, the population of interest may have long runs of homozygosity (ROH) due to recent inbreeding or/and small population size. This can affect the relatedness estimates. Here, we present a Hidden Markov Model (HMM) to estimate identity by descent (IBD) fragments and degree of relatedness up to 3rd degree, while additionally differentiating between siblings and parent-child with as low as 0.05x coverage DNA data. The output can be further refined with another HMM that we developed to estimate ROH proportions. We focus on the application of KIN on ancient human DNA, but it is applicable to other diploid species as well. On a machine with 8 cores, the generation of input files from 17 bam files takes -- minutes, and running KIN software takes -- minutes.
\end{abstract}

@@

However, it is not possible to directly observe IBD segments, and so a common approach is to first identify segments of the genome that are Identical by State (IBS) and using population allele frequencies obtained from a reference panel to calculate the probability of IBD given IBS \cite{vai_KINship_2020, boehnke_accurate_1997,lynch_estimation_1999, albrechtsen_natural_2010, purcell_plink_2007}.
\begin{figure}[h!]
    \includegraphics[width=18cm]{plots/inkscape_finalImg/schematic_sib.png}
    \centering
    \caption{IBD sharing between siblings without and with ROH. In top row, recombined chromosomes inherited from parents by a pair of siblings are shown. On the left, the parents have unrelated chromosomes, and on the right one parent has a homozygous pair of chromosomes. The bottom row shows the expected proportion of differences (p) along the chromosome in both the cases. p can take values of $p_0$, $p_1$ or $p_2$ which are expected proportion of differences in IBD states 0,1 and 2 respectively.  }
    \label{fig0:schematic}
\end{figure}

These methods utilize population allele frequencies, phase information, recombination maps, or genotype calls to co-estimate IBD and the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,li_accurate_2014,thornton_estimating_2012}.  PLINK, a popular tool to infer relatedness coefficient in diploid genotype data, estimates IBD probabilities from observed IBS, using the allele frequencies at each Single Nucleotide Polymorphism (SNP) in the analyzed samples \cite{purcell_plink_2007}. KING is another widely used software that allows relatedness inference with genotype data in homogeneous populations using population allele frequencies in the samples (KING homo), but can also work with structured populations with no good population allele frequencies using heterozygosity estimates from each sample (KING robust) \cite{manichaikul_robust_2010}.

@@

\subsection{Approaches that address problems with ancient DNA data}
One major issue with applying the above mentioned methods to ancient DNA data is that the sequence coverage is typically low, making it difficult to get genotype calls.
Several methods surmount this problem by using genotype likelihoods \cite{lipatov_maximum_2015,korneliussen_ngsrelate_2015}. In this way, it is possible to account for the uncertainty in genotype calls by summing over all possible genotypes weighted by their genotype likelihoods. However, these approaches typically still require at least 2x coverage, since genotype likelihoods may be imprecise at lower coverages . Ancient DNA analyses often face additional challenges such as contamination from present-day DNA \cite{peyregne_authentict_2020}, absence of suitable reference panels, and ascertainment bias in the case of data generated from DNA capture.

Several methods have been proposed to estimate relatedness without a reference panel, but many of these require either $>4$x coverage \cite{waples_allele_2019}, or a large sample size to get an estimate of population allele frequencies from the samples \cite{theunert_joint_2017}. In this study, we show that contamination can cause some of the individuals in ancient populations to look more unrelated than they are. This happens because the DNA sequence of an ancient individual typically exhibits more differences to a contaminant sequence than that of another ancient individual from the same population. It is worth mentioning that different DNA libraries may have different levels of contamination, even when they have been sampled from the same specimen. In addition, some analyzed  genomes  may have long runs of homozygosity (ROH), for example due to a small population size, or due to recent inbreeding. Long ROH cause related individuals to seem genetically more similar to each other, but do not affect the genetic distance between unrelated individuals.

Moreover, ancient DNA is commonly captured with a SNP array to enrich for sequences from the target species. Particularly methods based  on the fraction of sites in different IBS states are sensiteive to the ascertainment bias caused by this non-random selection of targted sites \cite{waples_allele_2019}.

Several limitations with ancient DNA mentioned above, such as low coverage, the absence of population allele frequencies or a reference panel, and ascertainment bias are resolved by a software called READ (REF). This software addresses the problem of unavailability of genotype calls by randomly sampling alleles from each individual. A string of these alleles at each position (called pseudo-haploids) are then compared to other individuals to calculate average pairwise genetic distances, which in turn are used to infer relatedness. However, READ only infers the degree of relatedness, and only up to second degree.

@@

The goal of KIN-HMM is to infer how two individuals are  relateded via the patterns of shared IBD-states along a pair of genomes. For this purpose,  we subdivide the genomes of a pair of individuals into $L$ large genomic windows (typically of size 10Mb), and infer the pattern of IBD-sharing for each relatedness case we consider (Unrelated, $5^{th}$ Degree, $4^{th}$ Degree, $3^{th}$ Degree, Grandparent-Grandchild, Avuncular, Half-siblings, Parent-Child, Siblings, and Identical). We then compare the likelihood between all models, and classify each pair to the model with the highest likelihood. We also return the most likely location of IBD tracts (using standard Viterbi algorithm), and the state posterior probablities $\BZ$.

The details of the likelihood computation are given in section \ref{ll}. As any HMM, KIN-HMM  require a set of n emission (section \ref{B}) and transition probability matrices. We assume the transition matrix for each relatedness case is fixed. Our emissions include a vector of parameters ($\delta$) that describes the variance in the data for each IBD state that we infer from the data (section ADD REF).

@@

$D_w$ is the number of sites that are different between a pair of individuals. For high-coverage data, this can be estimated accurately by comparing genotypes, but for low-coverage samples, $D_w$ needs to be estimated. The simplest approach is is to  randomly-sampled a read from each position ("pseudo-haploid" sequences) \cite{haak_massive_2015}. However, such an approach may result in loss of data, and hence we calculate $D_w$ by implicitly summing over all possible samplings:
\begin{align}\label{eq:x}
D_w &= \sum_{s=1}^{N_w} \nu_i(s) (1-\nu_j(s)) + (1-\nu_i(s)) \nu_j(s)
\end{align}
Here, $\nu_i(s)$ and $\nu_j(s)$ are the proportions of reads carrying the derived allele at SNP index $s$ for individuals i and j respectively.
Throughout, we will use bold-face notation to refer to the vector (or matrix) collecting  all the terms, e.g. $\BD = (D_1, D_2, \dots D_L)$.

\subsection{Log likelihood of the KIN-HMM}\label{ll}
The KIN-HMM uses $\BD$ and $\BN$ to classify each window into three hidden states $Z_w \in$ ($0$, $1$, $2$), reflecting zero, one or two shared chromosomes IBD, respectively. To take ROH into account, we define the variable $H_w \in$ ($0$, $1$, $2$) that designates that zero, one or both individuals are homozygous in window $w$. Since $H_w$ is unobserved, in practice we use the estimates from ROH-HMM $h_{wj} = P(H_w = j)$.

There are three additional model parameters, $\pi, \mathbf{A}$ and $\delta$. The initial probabilities $\pi$ give the probabilities of being in each state $Z$ at the beginning of each chromosome, which we set to uniform for simplicity. The transition matrix $\mathbf{A}$ gives the probability of moving from state $i$ to state $j$, given by $a_{ij}$, and is fixed and estimated from simulations for each relatedness case. The overdispersion-parameter $\delta$ takes into account that SNPs in each window vary in their allele frequency (see next setion).  For compactness of notation, we group the fixed parameters: $\theta = (\BN, \mathbf{A}, \pi)$.

@@

The proportion of differences between unrelated individuals $p_0$ is an important parameter. We follow READ \cite{kuhn_estimating_2018} and estimate $p_0$ as the median of differences for all possible pairs of individuals, which should work well if the majority of individuals in the samples are unrelated. We set $p_1$ to $(3/4)p_0$ and $p_2$ to $p_0/2$, because one, respectively two of the four compared chromosomes are identical in these cases.

@@

\paragraph{Initialization}
The value of $\delta_i$ is unknown to start with, and we set it to a random value between x and y.

@@

\paragraph{Maximization step}
The only free parameters we estimate in the M-step are the overdispersion parameters $\delta_i$. We do this optimization using a cost function, which is the log-emission probability weighted by the posterior probabilities of the hidden states $\gamma_{wi}$ and optionally the ROH state-probabilities $h_{w\omega}$ obtained from the ROH-HMM.

\begin{align}\label{eq:4}
\mathcal{C} &= \mathbb{E}[\log P(D_w|Z_w, H_w, \theta, \delta^{(t-1)}]\nonumber\\
&= \sum_{w=1}^L \sum_{i=0}^2\sum^2_{\omega=0} \log P(D_{w}|Z_w, H_w, N_w, \delta_{i \omega}^{(t-1)}, p_{i \omega})h_{w\omega}\gamma_{wi}
\end{align}

@@

The cost function \ref{eq:6} has one independent term for each $i$ and so we can separate them and estimate each $\delta_i$ independently using the XXX algorithm implemented in \texttt{scipy.optimize} (REF scipy).

@@

Unconstrained optimization of the $\delta_i$ could result in some confounding of cases, We realized that different cases of relatedness have different number of IBD states possible. For example, siblings may have all three IBD states present while a Parent-Child has only $Z_w=1$. However, the parent-child model could fit data generated under the sibling model by assigning it a very high $\delta_1$, which would reduce performance  \ref{figS4:bndsbeta}. We avoid this problem by constraining the $\delta$ such that the beta distributions for the different cases overlap by at most one standard deviation.

\subsection{Model comparison}\label{model_comp}
To infer the most probable relatedness case, we run our model on all cases ( section \ref{method_overview}) and compare the resulting likelihoods. We output the relatedness corresponding to the maximum likelihood model, and the confidence as given by the log likelihood ratio between the two highest likelihood models. The log likelihood-ratio is commonly used in likelihood-ratio tests to compare nested models. However, in our cases, models are not nested, and so standard likelihood-ratio theory does not apply. For this reason, we use the log likelihood ratio between the two best models as a statistic to assess the confidence in our classifications, and use simulations to obtain critical values (Fig \ref{fig3:cutoff}).

@@

\subsection{ROH estimation model}\label{roh}
Our HMM to detect ROH tracts works similar to the KIN-HMM described above, but we only consider one individual at a time, and only positions covered by at least two reads. For each site, we calculate the proportion of reads that carry different alleles, and sum them up in windows along the genome. We call the vector with the number of differences $\mathbf{\Delta}$, and the vector with the number of sites with at least two reads $\mathbf{N}$. Our model has two possible hidden states: homozygous state ($Y_w=3$), and non homozygous state ($Y-w=2$). As above, we collect the hidden states in a vector $\mathbf{Y} = (Y_1, \dots Y_w, \dots, Y_L)$. The complete data likelihood for the model in this case is then:

\begin{align}\label{eq:10}
    log P(\mathbf{\Delta},\mathbf{Y}|\Theta) &= log P(\mathbf{\Delta}|\mathbf{Y},\Theta) + log P(\mathbf{Y}|\Theta)\nonumber\\
 &= [\sum_{w} log P(\Delta_w|Y_w, \Theta) + \sum_{w} log P(Y_w|Y_w-1, \Theta)] + P(Y_0| \Theta)
\end{align}

@@

\begin{align}\label{eq:11}
\mathcal{C} = \sum_{w=1}^L \sum_{i=2}^3 P(\Delta_w|\delta_{i},p_{i}) \Gamma_{wi} ,
\end{align}
where $P(\Delta_w|\delta_{i},p_{i})$ is a beta-binomial probability with mean $p$ and overdispersion parameter $\eta$ similar to eq. \ref{eq:3}.

To estimate transitions, we initialize the transition matrix with 0.2 for the off-diagonal entries, and update it using the standard Baum-Welch update step \cite{baum_maximization_1970}.

@@

\subsection{Contamination correction}\label{contam}
Contamination by the DNA from present-day people is a common feature in human ancient DNA \cite{peyregne_present-day_2020}. To address this issue, we develop a heuristic that adjusts both $D_w$ and $N_w$ to remove contamination from the data.

@@

Then, by linearity of expectation, we obtain our estimators for the expected number of endogenous comparisons with a difference as 
\begin{align}
    \mathbb{E}[D_e] &= D_{obs} P(E|D) = D_{obs} \times \frac{P(D|E)P(E)}{P(D|E)P(E) + P(D| \neg E)P( \neg E)}
\end{align}.

@@

\begin{align}
    P(D|E) = p_e = \frac{\frac{\sum_w D_w}{\sum_w N_w} - C_{ij} \phi}{1 - C_{ij}}.
\end{align}

@@

\begin{equation}
 \mathbb{E}[D_e]= D_{obs}\frac{p_e (1-C_{ij})}{p_e(1-C_{ij}) + C_{ij}\phi}
 \end{equation}

@@

 \begin{equation}
\mathbb{E}[S_e] = S_{obs} P(E| \neg D) = S_{obs} \frac{(1-p_e)(1-C_{ij})}{(1-p_e)(1-C_{ij}) + C_{ij}(1-\phi)}.
 \end{equation}

Finally, we set $\mathbb{E}[N_e] = \mathbb{E}[S_e] + \mathbb{E}[D_e]$

@@

\paragraph{Critical Values}
 To investigate the limits of our method, we plotted the true-positive and false-positive rates when we use a particular difference in log-likelihood as a cutoff( Fig.\ref{figS10:cutoff},  Fig.\ref{fig4:Comparison_READ_KIN}).   Fig.\ref{figS10:cutoff} shows that for all relatedness cases except for $3^{rd}$-degree, the false positive rate is below $5\%$ even when simply selecting the model with the highest likelihood. We observe that for $3^{rd}$-degree relatives, using a cutoff of 1.0 brings down the false positive rate close to $5\%$ for all coverages except 0.05x. Thus, we  recommended to use a cutoff of 1 for all cases where ROH or contamination are not a concern.

\paragraph{Evaluating the ROH detection}
We next evaluate the performance of the ROH-HMM( Fig.\ref{fig2:ROH}). We observe that as coverage decreases, the raw heterozygosity estimates become more noisy. In the case without ROH (Fig \ref{fig2:ROH}A), we correctly infer that there is no ROH for all but two windows. We observe that the model predicts ROH probabilities well in all cases, even at a coverage of 0.1x \ref{tab:Table 2}.

@@

\paragraph{Comparing KIN-HMM with READ}
In Figure Fig. \ref{fig4:Comparison_READ_KIN}, we compare the classification accuracy of KIN (cutoff: 1 log-likelihood unit) to that of READ (cutoff: 1 standard deviation)  under four different simulation scenarios. We show that both methods have similar performance for low-coverage shotgun data (``control''-case),  except for $2^{nd}$-degree relatives, where KIN has higher power, especially at low coverages. We show that adding contamination adversely affects the True Positive rate for $2^{nd}$ Degree, and both the False Positve and True Positive rates corresponding to 2nd Degree relations for READ, while this does not affect KIN performance too much.

@@

\paragraph{Evaluating IBD prediction}
KIN-HMM co-estimates IBD states and relatedness, and hence it has a higher power to classify $2^{nd}$ Degree relatives, and can distinguish between Siblings and Parent-Child. We investigate our limit to predict IBD states along the genome by calculating the number of genomic windows where we correctly  predict IBD state for a pair of individuals. We calculate mean and standard deviation for this accuracy over 60 simulations of a pedigree (see section \ref{simulat}). Fig. \ref{fig4:IBDstate_accu} shows that we can predict IBD states for Unrelated $3^{rd}$ Degree, $2^nd Degree$, and Sibling pairs very well at high coverages, and our accuracy decreases with downsampling. On the other hand, our accuracy measure for Identical individuals and Parent-Child remains very high, due to robust relatedness estimates for these cases even at 0.03 x.

@@

\subsubsection{Application to Chagyrskaya Neanderthals}
To test KIN on real ancient dataset, we applied it to a Neandertal data set  from Chagyrskaya cave in Siberia \cite{kolobova_archaeological_2020-1}, Russia. This dataset contains genetic data from  14 fossils that might  possibly belong to contemporary Neandertals who occupied the cave 63 ± 4 to 48 ± 3 kya \cite{laurits_skov_genetic_nodate}. The data has low-to intermediate coverage ranging from 0.01x to 12.34x, with 8 samples at  $<1x$ coverage. Some of these specimens showed signs of long ROH, and contamination from modern humans as well as Hyena \cite{laurits_skov_genetic_nodate}. All these samples were captured with an array targeting for variable sites in high coverage Neandertal and Denisovan genomes and common variations in Africans \cite{laurits_skov_genetic_nodate}, and we further restricted  this data to the variable sites in two high-coverage Neandertal genomes: Altai Neandertal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017}. Our results for pairwise relatedness for these individuals is shown in Fig.\ref{fig5:Chagyrskaya_KIN}. We found three fossils from the same  individual (Chagyrskaya13-Chagyrskaya19-Chagyrskaya1141), a parent-child pair  (Chagyrskaya07 and Chagyrskaya17), and a pair of  $2^{nd}$-degree relatives (Chagyrskaya01/Chagyrskaya60). We also find that Chagyrskaya 06 and Chagyrskaya 14 are classified as parent-child with low confidence, but from the morphology they likely stem from the same individual. We believe that the contradiction is due to non human contamination present in these libraries ($1\%$ and $2.9\%$ respectively). Further, we identify   Chagyrskaya 07 and Chagyrskaya 60 as 3rd-degree relatives.

>>>>

\title{KIN: A method to infer relatedness from low-coverage ancient DNA}
\author{Divya Ratan Popli, Stéphane Peyrégne, Benjamin M. Peter}
%\date{5 August 2021}
\linenumbers

@@

\noindent Inferring familial relationships is an important question in population genetics. Knowing genetic kinship of ancient individuals leads to insights into the culture and social hierarchy of ancient populations, and is  relevant for downstream genetic analyses. However, estimating relatedness from ancient DNA is difficult since DNA sequences extracted from ancient remains may have low coverage, ascertainment bias, or contamination from various sources.

Here, we present KIN, a Hidden-Markov-Model-based approach to estimate identity-by descent-fragments and degree of relatedness up to 3rd-degree, from as low as 0.05x coverage ancient DNA data, and additionally differentiates between siblings and parent-child relationships. KIN also incorporates an adjustment for contamination, and an additional model to estimate the location of long runs of homozygosity, and uses it to improve classification accuracy.

We show in simulations that KIN performs at near perfect classification accuracy at inferring first-degree relationships, and at 90\% accuracy for second degree even at 0.05x coverage.

We illustrate the utility of KIN on two applications, one on a Neandertal population from Siberia, Russia, and one in a Bronze-Age population from the Lech Valley, Germany. 
We find that in both cases, KIN is competitive with existing approaches and provides new insights.

Thus, KIN is a useful tool for analyzing relatedness for large ancient-DNA datasets with low-coverage sequencing data that is applicable to any diploid species, without the requirement for a reference panel.
\end{abstract}

@@

However, it is not possible to directly observe IBD segments, and so a common approach is to first identify segments of the genome that are Identical by State (IBS) and using population allele frequencies obtained from a reference panel to calculate the probability of IBD given IBS. There are several methods that utilize population allele frequencies, phase information, recombination maps, or genotype calls to co-estimate IBD and the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,li_accurate_2014,thornton_estimating_2012, boehnke_accurate_1997,lynch_estimation_1999, albrechtsen_natural_2010, purcell_plink_2007,manichaikul_robust_2010}.

\begin{figure}[h!]
    \includegraphics[width=18cm]{plots/inkscape_finalImg/schematic_sib.png}
    \centering
    \caption{IBD sharing between siblings without and with ROH. In top row, recombined chromosomes inherited from parents by a pair of siblings are shown. On the left, the parents have unrelated chromosomes, and on the right one parent has a homozygous pair of chromosomes. The bottom row shows the expected proportion of differences (p) along the chromosome in both the cases. p can take values of $p_0$, $p_1$ or $p_2$ which are expected proportion of differences in IBD states 0,1 and 2 respectively.  }
    \label{fig0:schematic}
\end{figure}

@@

\subsection{Approaches that address problems with ancient DNA data}
One major issue with applying the above mentioned methods to ancient DNA data is that the sequence coverage is typically low, making it difficult to get genotype calls.
Several methods surmount this problem by using genotype likelihoods \cite{lipatov_maximum_2015,korneliussen_ngsrelate_2015}. In this way, it is possible to account for the uncertainty in genotype calls by summing over all possible genotypes weighted by their genotype likelihoods. However, these approaches typically still require at least 2x coverage, since genotype likelihoods may be imprecise at lower coverages \cite{korneliussen_angsd_2014} . Ancient DNA analyses often face additional challenges such as contamination from present-day DNA \cite{peyregne_present-day_2020-1}, absence of suitable reference panels, and ascertainment bias in the case of data generated from DNA capture \cite{prufer_computational_2010, vai_kinship_2020}.

Several methods have been proposed to estimate relatedness without a reference panel, but many of these require either $>4$x coverage \cite{waples_allele_2019}, or a large sample size to get an estimate of population allele frequencies from the samples \cite{theunert_joint_2017}. A second issue is contamination. If the contamination stems from another population, contaminated data will look more dissimilar to other individuals from the analyzed population, and hence relatedness inference will be biased towards towards less relatedness.  In addition, some analyzed  genomes  may have long runs of homozygosity (ROH), for example due to a small population size, or due to recent inbreeding. Long ROH cause related individuals to seem genetically more similar to each other, but do not affect the genetic distance between unrelated individuals.

Moreover, ancient DNA is commonly captured with a SNP array to enrich for sequences from the target species. Particularly methods based  on the fraction of sites in different IBS states are sensitive to the ascertainment bias caused by this non-random selection of targted sites \cite{waples_allele_2019}.

Several limitations with ancient DNA mentioned above, such as low coverage, the absence of population allele frequencies or a reference panel, and ascertainment bias are resolved by a software called READ \cite{kuhn_estimating_2018}. This software addresses the problem of unavailability of genotype calls by randomly sampling alleles from each individual. A string of these alleles at each position (called pseudo-haploids) are then compared to other individuals to calculate average pairwise genetic distances, which in turn are used to infer relatedness. However, READ only infers the degree of relatedness, and only up to second degree.

@@

The goal of KIN-HMM is to infer how two individuals are related via the patterns of shared IBD-states along a pair of genomes. For this purpose,  we subdivide the genomes of a pair of individuals into $L$ large genomic windows (typically of size 10Mb), and infer the pattern of IBD-sharing for each relatedness case we consider (Unrelated, $5^{th}$ Degree, $4^{th}$ Degree, $3^{th}$ Degree, Grandparent-Grandchild, Avuncular, Half-siblings, Parent-Child, Siblings, and Identical). We then compare the likelihood between all models, and classify each pair to the model with the highest likelihood. We also return the most likely location of IBD tracts (using standard Viterbi algorithm), and the state posterior probablities $\BZ$.

The details of the likelihood computation are given in section \ref{ll}. As any HMM, KIN-HMM requires a set of emission (section \ref{B}) and transition probability matrices. We assume the transition matrix for each relatedness case is fixed. Our emissions include a vector of parameters ($\delta$) that describe the variance in the data for each IBD state that we infer from the data (section \ref{B}).

@@

$D_w$ is the number of sites that are different between a pair of individuals. For high-coverage data, this can be estimated accurately by comparing genotypes, but for low-coverage samples, $D_w$ needs to be estimated. The simplest approach is is to  randomly-sample a read from each position to make "pseudo-haploid" sequences \cite{haak_massive_2015}. However, such an approach may result in loss of data, and hence we calculate $D_w$ by implicitly summing over all possible samplings:
\begin{align}\label{eq:x}
D_w &= \sum_{s=1}^{N_w} \nu_i(s) (1-\nu_j(s)) + (1-\nu_i(s)) \nu_j(s)
\end{align}
Here, $\nu_i(s)$ and $\nu_j(s)$ are the proportions of reads carrying the derived allele at SNP index $s$ for individuals i and j respectively.
Throughout, we will use bold-face notation to refer to the vector (or matrix) collecting  all the terms, e.g. $\BD = (D_1, D_2, \dots D_L)$.

\subsection{Log likelihood of the KIN-HMM}\label{ll}
The KIN-HMM uses $\BD$ and $\BN$ to classify each window into three hidden states $Z_w \in$ ($0$, $1$, $2$), reflecting zero, one or two shared chromosomes IBD, respectively. To take ROH into account, we define the variable $H_w \in$ ($0$, $1$, $2$) that designates that zero, one or both individuals are homozygous in window $w$. Since $H_w$ is unobserved, in practice we use the estimates from ROH-HMM: $h_{wj} = P(H_w = j)$.

There are three additional model parameters, $\pi, \mathbf{A}$ and $\delta$. The initial probabilities $\pi$ give the probabilities of being in each state $Z_0$ (at the beginning of each chromosome), which we set to uniform for simplicity. The transition matrix $\mathbf{A}$ gives the probability of moving from state $i$ to state $j$, given by $a_{ij}$, and is fixed and estimated from simulations for each relatedness case. The overdispersion-parameter $\delta$ takes into account that SNPs in each window vary in their allele frequency (see next setion).  For compactness of notation, we group the fixed parameters: $\theta = (\BN, \mathbf{A}, \pi)$.

@@

The proportion of differences between unrelated individuals $p_0$ is an important parameter. We follow READ \cite{kuhn_estimating_2018} and estimate $p_0$ as the median of differences for all possible pairs of individuals, which should work well if the majority of individuals in the samples are unrelated.

@@

\paragraph{Initialization}
The value of $\delta_i$ is unknown to start with, and we set it to a random value between 0 and 1000.

@@

\paragraph{Maximization step}
The only free parameters we estimate in the M-step are the overdispersion parameters $\delta_i$. We do this optimization using a cost function, which is the log-emission probability weighted by the posterior probabilities of the hidden states $\gamma_{wj}$ and optionally the ROH state-probabilities $h_{w\omega}$ obtained from the ROH-HMM.

\begin{align}\label{eq:4}
\mathcal{C} &= \mathbb{E}[\log P(D_w|Z_w, H_w, \theta, \delta^{(t-1)}]\nonumber\\
&= \sum_{w=1}^L \sum_{j=0}^2\sum^2_{\omega=0} \log P(D_{w}|Z_w, H_w, N_w, \delta_{j \omega}^{(t-1)}, p_{j \omega})h_{w\omega}\gamma_{wj}
\end{align}

@@

The cost function \ref{eq:6} has one independent term for each $i$ and so we can separate them and estimate each $\delta_i$ independently using the minimize\_scalar algorithm implemented in \texttt{scipy.optimize} \cite{virtanen_scipy_2020}.

@@

Unconstrained optimization of the $\delta_i$ could result in some confounding of cases, We realized that different cases of relatedness have different number of IBD states possible. For example, siblings may have all three IBD states present while a Parent-Child has only $Z_w=1$. However, the parent-child model could fit data generated under the sibling model by assigning it a very high $\delta_1$, which would reduce performance  (for example, see fig. S \ref{figS4:bndsbeta}). We avoid this problem by constraining the $\delta$ such that the beta distributions for the different cases overlap by at most one standard deviation.

\subsection{Model comparison}\label{model_comp}
To infer the most probable relatedness case, we run our model on all cases (section \ref{method_overview}) and compare the resulting likelihoods. We output the relatedness corresponding to the maximum likelihood model, and the confidence as given by the log likelihood ratio between the two highest likelihood models. The log likelihood-ratio is commonly used in likelihood-ratio tests to compare nested models. However, in our case, models are not nested, and so standard likelihood-ratio theory does not apply. For this reason, we use the log likelihood ratio between the two best models as a statistic to assess the confidence in our classifications, and use simulations to obtain critical values (Fig S \ref{figS10:cutoff}).

@@

\subsection{ROH estimation model}\label{roh}
Our HMM to detect ROH tracts works similar to the KIN-HMM described above, but we only consider one individual at a time, and only positions covered by at least two reads. For each site, we calculate the proportion of reads that carry different alleles, and sum them up in windows along the genome. We call the vector with the number of differences $\mathbf{\Delta}$, and the vector with the number of sites with at least two reads $\mathbf{N}$. Our model has two possible hidden states: homozygous state ($Y_w=3$), and non homozygous state ($Y_w=2$). As above, we collect the hidden states in a vector $\mathbf{Y} = (Y_1, \dots Y_w, \dots, Y_L)$. The complete data likelihood for the model in this case is then:

\begin{align}\label{eq:10}
    log P(\mathbf{\Delta},\mathbf{Y}|\Theta,\delta) &= log P(\mathbf{\Delta}|\mathbf{Y},\Theta, \delta) + log P(\mathbf{Y}|\Theta)\nonumber\\
 &= [\sum_{w} log P(\Delta_w|Y_w, \Theta, \delta) + \sum_{w} log P(Y_w|Y_{w-1}, \Theta)] + P(Y_0| \Theta)
\end{align}

@@

\begin{align}\label{eq:11}
\mathcal{C} = \sum_{w=1}^L \sum_{i=2}^3 P(\Delta_w|\delta_{i},p_{i}) \Gamma_{wi} ,
\end{align}
where $P(\Delta_w|\delta_{i},p_{i})$ is a beta-binomial probability with mean $p$ and overdispersion parameter $\delta$ similar to eq. \ref{eq:3}.

To estimate transitions, we initialize the transition matrix with the value 0.2 for the off-diagonal entries, and update it using the standard Baum-Welch update step \cite{baum_maximization_1970}.

@@

\subsection{Contamination correction}\label{contam}
Contamination by the DNA from present-day people is a common feature in human ancient DNA \cite{peyregne_present-day_2020}. To address this issue, we developed a heuristic that adjusts both $D_w$ and $N_w$ to remove contamination from the data.

@@

Then, by linearity of expectation, we obtain our estimator for the expected number of endogenous comparisons with a difference as 
\begin{align}
    \mathbb{E}[D_w^{'}] &= D_{w} P(E|D) = D_{w} \times \frac{P(D|E)P(E)}{P(D|E)P(E) + P(D| \neg E)P( \neg E)}
\end{align}.

@@

\begin{align}
    P(D|E) = \rho = \frac{\frac{\sum_w D_w}{\sum_w N_w} - C_{ij} \phi}{1 - C_{ij}}.
\end{align}

@@

\begin{equation}
 \mathbb{E}[D_w^{'}]= D_{w}\frac{\rho (1-C_{ij})}{\rho(1-C_{ij}) + C_{ij}\phi}
 \end{equation}

@@

 \begin{equation}
\mathbb{E}[S_w^{'}] = S_{w} P(E| \neg D) = S_{w} \frac{(1-\rho)(1-C_{ij})}{(1-\rho)(1-C_{ij}) + C_{ij}(1-\phi)}.
 \end{equation}

Here, $S_w = N_w - D_w$. Hence, we set $\mathbb{E}[N_w^{'}] = \mathbb{E}[S_w^{'}] + \mathbb{E}[D_w^{'}]$. We do a similar contamination correction for the input of ROH-HMM.

@@

\paragraph{Critical Values}
 To investigate the limits of our method, we plotted the true-positive and false-positive rates when we use a particular difference in log-likelihood as a cutoff( Fig.\ref{figS10:cutoff},  Fig.\ref{fig3:Comparison_READ_KIN}).   Fig.\ref{figS10:cutoff} shows that for all relatedness cases except for $3^{rd}$-degree, the false positive rate is below $5\%$ even when simply selecting the model with the highest likelihood. We observe that for $3^{rd}$-degree relatives, using a cutoff of 1.0 brings down the false positive rate close to $5\%$ for all coverages except 0.05x. Thus, we  recommended to use a cutoff of 1 for all cases where ROH or contamination are not a concern.

\paragraph{Evaluating the ROH detection}
We next evaluate the performance of the ROH-HMM( Fig.\ref{fig2:ROH}). We observe that as coverage decreases, the raw heterozygosity estimates become more noisy. In the case without ROH (Fig \ref{fig2:ROH}A), we correctly infer that there is no ROH for all but two windows. We observe that the model predicts ROH probabilities well in all cases, even at a coverage of 0.1x (Table \ref{tab:Table 2}).

@@

\paragraph{Comparing KIN-HMM with READ}
In Figure Fig. \ref{fig3:Comparison_READ_KIN}, we compare the classification accuracy of KIN (cutoff: 1 log-likelihood unit) to that of READ (cutoff: 1 standard deviation)  under four different simulation scenarios. We show that both methods have similar performance for low-coverage shotgun data (``control''-case),  and for ascertained data. One exception are  $2^{nd}$-degree relatives, where KIN has higher power, especially at low coverages. We also find that READ is strongly impacted by contamination, but that the correction implemented in KIN is sufficient to remove this bias.. For simulations with ROH, we find the overall classification accuracy is only slightly reduced. Since ROH makes individuals more similar, we see a bias towards more false positives, for first and second-degree relationships, which is slightly higher for READ than for KIN.

@@

\paragraph{Evaluating IBD prediction}
KIN-HMM co-estimates IBD states and relatedness, and hence it has a higher power to classify $2^{nd}$ Degree relatives, and can distinguish between Siblings and Parent-Child. We investigate our limit to predict IBD states along the genome by calculating the number of genomic windows where we correctly  predict IBD state for a pair of individuals. We calculate mean and standard deviation for this accuracy over 60 simulations of a pedigree (see section \ref{simulat}). Fig. \ref{fig4:IBDstate_accuracy} shows that we can predict IBD states for Unrelated, $3^{rd}$ Degree, $2^{nd} Degree$, and Sibling pairs very well at high coverages, and our accuracy decreases with downsampling. On the other hand, our accuracy measure for Identical individuals and Parent-Child remains very high, due to robust relatedness estimates for these cases even at 0.03 x.

@@

\subsubsection{Application to Chagyrskaya Neanderthals}
To test KIN on real ancient dataset, we applied it to a Neandertal data set  from Chagyrskaya cave in Siberia, Russia \cite{kolobova_archaeological_2020-1, mafessoni_high-coverage_2020}. This dataset contains genetic data from  14 fossils that might  possibly belong to contemporary Neandertals who occupied the cave 63 ± 4 to 48 ± 3 kya \cite{laurits_skov_genetic_nodate}. The data has low-to-intermediate coverage ranging from 0.01x to 12.34x, with 8 samples at  $<1x$ coverage. Some of these specimens showed signs of long ROH, and contamination from modern humans as well as Hyena \cite{laurits_skov_genetic_nodate}. All these samples were captured with an array targeting for variable sites in high coverage Neandertal and Denisovan genomes and common variations in Africans \cite{laurits_skov_genetic_nodate}, and we further restricted  this data to the variable sites in two high-coverage Neandertal genomes: Altai Neandertal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017}. Our results for pairwise relatedness for these individuals is shown in Fig.\ref{fig5:Chagyrskaya_KIN}. We found three fossils from the same  individual (Chagyrskaya13-Chagyrskaya19-Chagyrskaya1141), a parent-child pair  (Chagyrskaya07 and Chagyrskaya17), and a pair of  $2^{nd}$-degree relatives (Chagyrskaya01/Chagyrskaya60). Further, we identify   Chagyrskaya 07 and Chagyrskaya 60 as 3rd-degree relatives.

Our estimates are consistent with those obtained using READ, except that READ is unable to detect 3rd-degree relationships, and cannot distinguish between siblings and parent-child. We also find that both  KIN and READ classify Chagyrskaya 06 and Chagyrskaya 14 as parent-child with low confidence, but from the morphology they likely stem from the same individual. We believe that this due to non-human contamination present in these libraries ($1\%$ and $2.9\%$ respectively).
